Exploring the linear separablity of syntactic and semantic information in BERT embeddings

### 2022-07-30 paper/paper.tex

A novel method for removing information from an embedding, INLP iteratively trains linear models on a specific classification task, and projects the input on intersection of the nullspaces of those linear models. Our objective is that, by applying the INLP procedure to a syntactic task, we are able to separate the representation into a syntactic space and a non-syntactic space. We then compare the performance of a model that seeks to classify semantic labels using the original BERT embeddings with an otherwise identical model trained on embeddings projected onto the non-syntactic space, in order to see if BERT is using syntactic information when performing the semantic task. Conversely, we can also first probe a semantic task, thus defining a semantic and non-semantic space, and then investigate the performance of embeddings projected onto those spaces when performing a syntactic classification task. Once we derive the semantic space and syntactic space from the experiment, we further investigate on the separability of the two spaces by comparing the embedding projections. %See section ???
%Make the research question very clear
%Should a model be able to solve a semantic task without syntactic information?
%Vice versa? (Just be very explicit here)

@@

The original dataset does not include CCG tags, however \citealp{pmbData} utilized a CCG parser to produce CCG tags. We follow a similar procedure and apply a CCG parser \citep{ccg_tagger} to develop word-level CCG tags. Once we obtain both CCG tags and semantic tags for the dataset, we can perform the syntactic and semantic probing task as desired.

@@

Similar to the above, the semantic probing task involves training a linear classifier on the final layer BERT embeddings in order to predict the semantic tag (described in the data section) associated with each word. This classifier is used in the INLP algorithm in order to create a guarding function, $P_{sem}$, for the information necessary to complete the semantic tag labeling task. As described in the syntactic probing task section, we use the resulting guarding function to decompose the original embedding into a ``non-semantic component'', $v_{no sem} = P_{sem} v_{orig}$, and a ``semantic component'', $v_{sem} = v_{orig} - v_{no sem}$.

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{p{4cm}p{3cm}}\hline
        \textbf{Expression} & \textbf{Description}\\ \hline 
        $v_{orig}$ & Original BERT embeddings  \\
        $v_{nosem} = P_{sem} v_{orig}$ & Gained after INLP with semantic task \\
        $v_{nosyn}= P_{syn} v_{orig}$ & Gained after INLP with syntactic task \\
        $v_{sem} = v_{orig}-v_{nosem}$ & Semantic representation \\
        $v_{syn} = v_{orig}-v_{nosyn}$ & Syntactic representation \\
        %$v_{syn-sem} = P_{sem} v_{syn}$ &Contains syntactic features only \\
        %$v_{sem-syn} = P_{syn} v_{sem}$ &Contains semantic features only \\
        % $v_{syn*sem} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{sem}\|}$ & Syntactic representations projected on semantic space\\
        % $v_{sem*syn} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{syn}\|}$ & Semantics representations projected on semantic space\\

@@

To ensure a fair assessment of the impact of the information loss, we conduct experiments for which we start with the original BERT embeddings and randomly remove the same number of directions that our derived embeddings lost. For the embeddings derived from projection alone, we record the number of removed directions using $Null(P)$, where $P$ is the projection matrix. For embeddings not derived from matrix multiplication, we obtain the number from $Null(M)$ where $M$ is the embedding matrix with size (768,instance number).\footnote{This is not necessarily equivalent to the number of direction removed. $Null(M) \geq Null(P)$ for the embedding matrix $M$ that corresponds to projection $P$, but in practice the numbers are very close. With this approach we err on the side of removing more random directions from our random embeddings, resulting in a more conservative comparison when assessing our derived embeddings.} Then we create embeddings with the same number of directions randomly removed and train the linear classifiers on these embeddings. The testing accuracies from our experiments can be found in table \ref{role description}.

@@

\section{Discussion}
\label{sec:discussion}
The INLP method successfully guards the information used in the probing task. When performing CCG tagging, the model accuracy drops significantly when comparing the $v_{nosyn}$ embeddings with the Rand($|v_{nosyn}|$) embeddings (from 82.26\% to 23.76\%), and similarly for semantic tagging (from 87.71\% to 17.28\%). In contrast, when the directions are randomly removed, the performance remains relatively the similar to the classifiers' performances on the original BERT embeddings for both tasks (84.75\% and 88.56\%, respectively).

@@

What is surprising is that the loss of semantic information has a more significant impact on the syntactic classification task than the loss of syntactic information does on the semantic tagging task. These results suggest that the semantics task is less dependent on syntactic information than the opposite direction, which challenges the assumption that syntactic information is more pertinent to semantic comprehension than vice versa.

@@

Using these guarding functions we were able to explore the importance and separability of the syntactic and semantic information contained in BERT embeddings. We evaluated the classification tasks on various derived embeddings and concluded that not only is the linear syntactic and semantic information essential for their respective classification tasks, these information sets are also very important for the opposing classification tasks as well. Thus the two information sets are not easily separated. Somewhat surprising, we determined that the semantic information was more influential on the success of the syntactic classifier than the other way around. 
%something about linear separability with last two embeddings.

@@

Though INLP successfully produces the desired result, it is worth noting that our dataset is relatively small compared to the number of parameters in linear classifier. In the future, we will expect this experiment to be reproduced at a larger scale so that the experiment findings can be further validated. Additionally, in future work we hope to perform a similar analysis on embeddings derived from different layers of the BERT architecture, in order to determine the importance and separability of these information sets at each layer. In doing so we hope to better understand how BERT processes different types of linguistic information throughout the encoding process.

>>>>

A novel method for removing information from an embedding, INLP iteratively trains linear models on a specific classification task, and projects the input on the intersection of the nullspaces of those linear models. Our objective is that, by applying the INLP procedure to a syntactic task, we are able to separate the representation into a syntactic space and a non-syntactic space. We then compare the performance of a model that seeks to classify semantic labels using the original BERT embeddings with an otherwise identical model trained on embeddings projected onto the non-syntactic space, in order to see if BERT is using syntactic information when performing the semantic task. Conversely, we can also first probe a semantic task, thus defining a semantic and non-semantic space, and then investigate the performance of embeddings projected onto those spaces when performing a syntactic classification task. The performance of these embeddings on their opposing classification tasks will also give us an indication of how linearly separable the two information sets are.%Once we derive the semantic space and syntactic space from the experiment, we further investigate on the separability of the two spaces by comparing the embedding projections. %See section ???
%Make the research question very clear
%Should a model be able to solve a semantic task without syntactic information?
%Vice versa? (Just be very explicit here)

@@

The original dataset does not include CCG tags, however \citealp{pmbData} utilized a CCG parser to produce CCG tags. We follow a similar procedure and apply a CCG parser \citep{ccg_tagger} to develop word-level CCG tags. Once we obtain both CCG tags and semantic tags for the dataset, we can perform the syntactic and semantic probing tasks as desired.

@@

Similar to the above, the semantic probing task involves training a linear classifier on the final layer BERT embeddings in order to predict the semantic tag associated with each word. This classifier is used in the INLP algorithm in order to create a guarding function, $P_{sem}$, for the information necessary to complete the semantic tag labeling task. As described in the syntactic probing task section, we use the resulting guarding function to decompose the original embedding into a ``non-semantic component'', $v_{no sem} = P_{sem} v_{orig}$, and a ``semantic component'', $v_{sem} = v_{orig} - v_{no sem}$.

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{p{4cm}p{3cm}}\hline
        \textbf{Expression} & \textbf{Description}\\ \hline 
        $v_{orig}$ & Original BERT embeddings  \\
        $v_{nosem} = P_{sem} v_{orig}$ & Gained after INLP with semantic task \\
        $v_{nosyn}= P_{syn} v_{orig}$ & Gained after INLP with syntactic task \\
        $v_{sem} = v_{orig}-v_{nosem}$ & Semantic representation \\
        $v_{syn} = v_{orig}-v_{nosyn}$ & Syntactic representation \\
        Rand($n$) & BERT embeddings with n random directions removed \\
        %$v_{syn-sem} = P_{sem} v_{syn}$ &Contains syntactic features only \\
        %$v_{sem-syn} = P_{syn} v_{sem}$ &Contains semantic features only \\
        % $v_{syn*sem} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{sem}\|}$ & Syntactic representations projected on semantic space\\
        % $v_{sem*syn} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{syn}\|}$ & Semantics representations projected on semantic space\\

@@

To ensure a fair assessment of the impact of the information loss, we conduct experiments for which we start with the original BERT embeddings and randomly remove the same number of directions that our derived embeddings lost. For the embeddings derived from projection matrices, we record the number of removed directions using $Null(P)$, where $P$ is the projection matrix. .\footnote{For embeddings not derived from matrix multiplication, we obtain the number from $Null(M)$ where $M$ is the embedding matrix with size (768,instance number). This is not necessarily equivalent to the number of direction removed. $Null(M) \geq Null(P)$ for the embedding matrix $M$ that corresponds to projection $P$, but in practice the numbers are very close. With this approach we err on the side of removing more random directions from our random embeddings, resulting in a more conservative comparison when assessing our derived embeddings. Also note that we decided not to include these types of embeddings in our final assessment due to concerning behavior we witnessed after performing the linear transformation to obtain the embeddings. Specifically, the rank of the resulting embedding matrix did not match our expectations. This may be due to a rounding or an issue with the orthonormalization process, but we were unable to confirm. So we decided to remove all embeddings derived in this manner from our evaluation process until we are able to determine the source of this behavior. This means that $v_{syn}$ and $v_{sem}$ will not be included in our final evaluation.} Then we create embeddings with the same number of directions randomly removed and train the linear classifiers on these embeddings. The testing accuracies from our experiments can be found in table \ref{role description}.

@@

\section{Discussion}
\label{sec:discussion}
The INLP method successfully guards the information used in the probing task. When performing CCG tagging, the model accuracy drops significantly when comparing the $v_{nosyn}$ embeddings with the Rand($|v_{nosyn}|$) embeddings (from 82.26\% to 23.76\%), and similarly for semantic tagging (from 87.71\% to 17.28\%). In contrast, when the directions are randomly removed, the performance remains relatively the similar to the classifiers' performances on the original BERT embeddings for both tasks (84.75\% and 88.56\%, respectively). This suggests that the syntactic and semantic information contained in the original BERT embeddings is not highly concentrated and that removing a small amount of one information set will not have a significant impact on the classification task.

@@

What is surprising is that the loss of semantic information has a more significant impact on the syntactic classification task than the loss of syntactic information does on the semantic tagging task. These results suggest that the semantic task is less dependent on syntactic information than the opposite direction, which challenges the assumption that syntactic information is more pertinent to semantic comprehension than vice versa.

@@

Using the INLP-derived guarding functions we were able to explore the importance and separability of the syntactic and semantic information contained in BERT embeddings. We evaluated the classification tasks on various derived embeddings and concluded that not only is the linear syntactic and semantic information essential for their respective classification tasks, these information sets are also very important for the opposing classification tasks as well. Thus the two information sets are not easily separated. Somewhat surprising, we determined that the semantic information was more influential on the success of the syntactic classifier than the other way around. 
%something about linear separability with last two embeddings.

@@

Though INLP successfully produces the desired result, it is worth noting that our dataset is relatively small compared to the number of parameters in the linear classifier. In the future, we will expect this experiment to be reproduced at a larger scale so that the experiment findings can be further validated. Additionally, we plan to increase the variety of training and evaluation tasks in order to develop a more general understanding of how syntactic and semantic information is encoded in BERT embeddings. Finally, we intend to perform a similar analysis on embeddings derived from different layers of the BERT architecture, in order to determine the importance and separability of these information sets at each layer. In doing so we hope to better understand how BERT processes different types of linguistic information throughout the encoding process.
